---
title: "Presentation of Data"
author: "ENGEN102-23G (HAM) - Engineering Maths and Modelling 1B"
format:
  revealjs: 
    slide-number: c/t
    progress: true
    auto-stretch: false
    height: 1080
    width: 1920
    theme: reveal-extra/quartz.scss
    template-partials:
      - reveal-extra/title-block.html
    highlight-style: breeze
    code-line-numbers: false
    embed-resources: false
editor: source
---

```{r setup}
#| include: false

library(ggplot2)
library(knitr)
```

## What is Data?

. . .

> In statistical parlance, **data** is a plural word referring to a collection of numbers or other pieces of information to which meaning has been attached.

::: {style="font-size: 0.9em; text-align: right; margin-top: -0.5em;"}
 --- Utts (2014)
:::

::: {.aside}
Utts, J. M. (2014). *Seeing Through Statistics* (4th ed.). Cengage Learning.
:::

## Observations[, values]{.fragment fragment-index=1}[, and variables]{.fragment fragment-index=2}

An [observation]{.tomato} is an individual that we measure *or* categorise **data** about.

<blockquote>
- An apple
- A hospital
- A cantilever beam
</blockquote>

:::: {.fragment fragment-index=1 style="margin-top: 50px;"}
A [value]{.steelblue} is a singular piece of **data** about an [observation]{.tomato}.

<blockquote style="border-left: 0.25em solid steelblue;">
- The weight of an apple
- The bed occupancy rate at Waikato Hospital
- The load at point [x]{.MathJax-Font} of a cantilever beam
</blockquote>
::::

:::: {.fragment fragment-index=2 style="margin-top: 50px;"}
A [variable]{.orchid} is a collection of one *type* of **data** collected on all [observations]{.tomato}.

<blockquote style="border-left: 0.25em solid orchid;">
- The weights of all apples
- The bed occupancy rates for all New Zealand hospitals
- The loads at point [x]{.MathJax-Font} for all cantilever beams
</blockquote>
::::

## More on values and variables

::::: {.columns}
:::: {.column .incremental}
**Numeric** <small>Quantitative</small>  
Data that can be described with "numbers".

<p>
- [Discrete]{.darkorange} <small>Integer</small>
  - Age (measured in years)
  - Number of siblings
</p>
<p>
- [Continuous]{.darkorange} <small>Real</small>
  - Systolic and diastolic blood pressure
  - 100 metre sprint time
</p>

::::
:::: {.column .fragment .incremental}
**Categorical** <small>Qualitative, Factor</small>  
Data that is best described with "text".

<p>
- [Nominal]{.darkorange} <small>No order</small>
  - Eye colour
  - Te Whatu Ora -- Health New Zealand regions^[**URL:** [careers.tewhatuora.govt.nz/locations/](https://careers.tewhatuora.govt.nz/locations/)]
</p>
<p>
- [Ordinal]{.darkorange} <small>Natural order</small>
  - Highest educational degree earned
  - Number of Wordle guesses
</p>

::::
:::::

## Preparing data for presentation

Let's start by (virtually) collecting data to estimate the *average* weight of these 100 blocks by selecting **only** 10 blocks.

[**URL:** [bit.ly/choc-and-blocks](https://bit.ly/choc-and-blocks)]{style="margin-left: 1em;"}

::: { style="margin-top: 50px;" .fragment fragment-index=0}
Note how the data about these 10 blocks was arranged. [Nowadays, this particular arrangement is known as the **tidy data** format (Wickham et al., 2023).]{.fragment fragment-index=1}
:::

::: {.incremental .fragment fragment-index=1}
<blockquote style="border-left: 0.25em solid #685f5f;">
1. Each [variable]{.orchid} is a column.
2. Each [observation]{.tomato} is a row.
3. Each [value]{.steelblue} is a cell.
</blockquote>
:::

::: {.aside}
Wickham, H., Çetinkaya-Rundel, M., & Grolemund, G. (2023). *R for Data Science* (2nd ed.). O'Reilly.
:::

## [Ex 1:]{.red} Compressive Strength of Al--Li Alloys

::::: {.columns}
:::: {.column width="70%"}
<blockquote style="border-left: 0.25em solid #685f5f;">
The `alloy.xlsx` file describes data about the compressive strength in 80 specimens of a new aluminium-lithium alloy undergoing evaluation as a possible material for aircraft structural elements.
</blockquote>

::: {.fragment fragment-index=1 .orchid-table style="font-size: 0.9em; margin-top: 50px; width: 90%;"}
+:-------------+:----------------------------------------------------------------------------------------------------------------------------+
| **strength** | A number that denotes the compressive strength of an aluminium-lithium alloy, measured in pounds per square inch (psi). 
+--------------+-----------------------------------------------------------------------------------------------------------------------------+
:::
::::
:::: {.column width="30%" .fragment fragment-index=0}
```{r alloy} 
#| results: asis
#| classes: excel

library(readxl)
alloy.df <- read_excel("datasets/alloy.xlsx")
 
rbind(colnames(alloy.df), head(alloy.df, n = 14), rep("...", ncol(alloy.df))) |>
  kable(format = "html", row.names = TRUE, col.names = LETTERS[1:ncol(alloy.df)])
```
::::
:::::

::: {.aside}
Data from Montgomery, D. C., Runger, G. C., & Hubele, N. F. (2010). *Engineering Statistics* (5th ed.). Wiley.
:::

## Stacked Dot Plot

A single numeric variable is plotted on a number line, where *similar* values are stacked on top of each other.

:::: {.r-stack style="margin-top: 50px;"}
::: {style="width: 50%; margin: auto; display: block;"}
```{r alloy-dotplot}
#| fig-dpi: 300
#| fig-height: 4
#| fig-width: 7
#| out-width: 100%
#| fig-cap: "**Figure:** Distribution of 80 aluminium-lithium alloy specimens' [**Figure:**]{style=\"visibility: hidden;\"} compressive strength."
plotObj <- ggplot(alloy.df, aes(x = strength)) +
  labs(x = "Compressive strength (psi)", title = "Distribution of Compressive Strength Al–Li Alloys") +
  theme_bw()

dotplot <- plotObj + 
  geom_dotplot(fill = NA, stroke = 3, colour = "black", binwidth = (max(alloy.df$strength) - min(alloy.df$strength)) / 25, dotsize = 0.7) + 
  labs(y = NULL) + 
  scale_y_continuous(breaks = NULL)

dotplot.xlims <- ggplot_build(dotplot)$layout$panel_scales_x[[1]]$range$range

plotObj + 
  geom_point(aes(y = 0.03), size = 4.5, stroke = 1.5, colour = "black", shape = 1) + 
  labs(y = NULL) + 
  coord_cartesian(dotplot.xlims,  ggplot_build(dotplot)$layout$panel_scales_y[[1]]$range$range, clip = "off") +
  scale_y_continuous(breaks = NULL)
```
:::
::: {.fragment style="width: 50%; margin: auto; display: block;"}
```{r alloy-stackeddotplot}
#| fig-dpi: 300
#| fig-height: 4
#| fig-width: 7
#| out-width: 100%
#| fig-cap: "[**Figure:** Distribution of 80 aluminium-lithium alloy specimens' **Figure:** compressive strength.]{style=\"visibility: hidden;\"}"

print(dotplot)
```
:::
::::

## Box Plot <small style="font-weight: normal !important;">Box-and-Whisker Plot</small>

A single numeric variable is sorted into **quarters**. The *central* 50% of the values are plotted as a segmented box, and the *remainder* are plotted as lines or [points]{.fragment fragment-index=2 .highlight-red}.

:::: {.r-stack style="margin-top: 50px;"}
::: {style="width: 50%; margin: auto; display: block;"}
```{r alloy-stackeddotplot}
#| fig-dpi: 300
#| fig-height: 4
#| fig-width: 7
#| out-width: 100%
#| fig-cap: "**Figure:** Distribution of 80 aluminium-lithium alloy specimens' [**Figure:**]{style=\"visibility: hidden;\"} compressive strength."
```
:::
::: {.fragment style="width: 50%; margin: auto; display: block;" fragment-index=0}
```{r alloy-stackeddotplot-annotated}
#| fig-dpi: 300
#| fig-height: 4
#| fig-width: 7
#| out-width: 100%
#| fig-cap: "[**Figure:** Distribution of 80 aluminium-lithium alloy specimens' **Figure:** compressive strength.]{style=\"visibility: hidden;\"}"

boxplot.stats <- ggplot_build(plotObj + geom_boxplot())$data[[1]]

super.dotplot <- plotObj + 
  geom_dotplot(fill = NA, stroke = 3, colour = "gray", binwidth = (max(alloy.df$strength) - min(alloy.df$strength)) / 25, dotsize = 0.7) + 
  labs(y = NULL) + 
  scale_y_continuous(breaks = NULL) +
  coord_cartesian(dotplot.xlims, clip = "off")

super.dotplot +
  geom_vline(xintercept = c(boxplot.stats$xmin_final, boxplot.stats$xlower, boxplot.stats$xmiddle, boxplot.stats$xupper, boxplot.stats$xmax_final), linetype = 5, linewidth = 1, colour = "black")
```
:::
::: {.fragment style="width: 50%; margin: auto; display: block;" fragment-index=1}
```{r alloy-boxplot}
#| fig-dpi: 300
#| fig-height: 4
#| fig-width: 7
#| out-width: 100%
#| fig-cap: "[**Figure:** Distribution of 80 aluminium-lithium alloy specimens' **Figure:** compressive strength.]{style=\"visibility: hidden;\"}"

plotObj + 
  geom_boxplot(colour = "black", outlier.stroke = 1.5, outlier.shape = 1, outlier.size = 4.5, outlier.colour = "black", fill = "lightblue") + 
  labs(y = NULL) +
  scale_y_continuous(breaks = NULL, limits = c(-1, 1)) +
  xlim(dotplot.xlims)
```
:::
::::

::: {.fragment fragment-index=2}

- The [points]{.red} are typically identified by using the *range* of the **whole** box
:::

## Histogram

A single numeric variable is *typically* sorted into **equally sized intervals**. Each interval is plotted as a bar, whose height is determined by the number of values *within* the interval.

:::: {.r-stack style="margin-top: 50px;"}
::: {style="width: 50%; margin: auto; display: block;"}
```{r alloy-stackeddotplot}
#| fig-dpi: 300
#| fig-height: 4
#| fig-width: 7
#| out-width: 100%
#| fig-cap: "**Figure:** Distribution of 80 aluminium-lithium alloy specimens' [**Figure:**]{style=\"visibility: hidden;\"} compressive strength."
```
:::
::: {.fragment style="width: 50%; margin: auto; display: block;" fragment-index=0}
```{r alloy-stackeddotplot-annotated-2}
#| fig-dpi: 300
#| fig-height: 4
#| fig-width: 7
#| out-width: 100%
#| fig-cap: "[**Figure:** Distribution of 80 aluminium-lithium alloy specimens' **Figure:** compressive strength.]{style=\"visibility: hidden;\"}"

histogram.stats <- ggplot_build(plotObj + geom_histogram(center = 0, binwidth = (max(alloy.df$strength) - min(alloy.df$strength)) / 12.5))$data[[1]]

super.dotplot +
  geom_vline(xintercept = c(histogram.stats$xmin, histogram.stats$xmax[nrow(histogram.stats)]), linetype = 5, linewidth = 1, colour = "black")
```
:::
::: {.fragment style="width: 50%; margin: auto; display: block;" fragment-index=1}
```{r alloy-histogram}
#| fig-dpi: 300
#| fig-height: 4
#| fig-width: 7
#| out-width: 100%
#| fig-cap: "[**Figure:** Distribution of 80 aluminium-lithium alloy specimens' **Figure:** compressive strength.]{style=\"visibility: hidden;\"}"

plotObj + 
  geom_histogram(center = 0, binwidth = (max(alloy.df$strength) - min(alloy.df$strength)) / 12.5, colour = "black", fill = "lightblue") +
  labs(y = NULL) +
  scale_y_continuous(breaks = NULL) +
  coord_cartesian(xlim = dotplot.xlims, clip = "off")
```
:::
::::

. . .

- An algorithm is often used to determine the number of intervals
- The intervals are usually right-closed *and* left-open

## Sample Mean

A *measure of centre*^[Some resources describe a *measure of centre* as a *measure of location* instead.] that is often coined as the balancing point of a numeric variable. The word **average** in everyday use is implicitly defined as the (sample) mean.

. . .

::: {.definition style="margin-top: 50px; width: 75%;"}
Let $x_i$ be the $i$th value of a numeric variable and $n$ be the total number of values. The *sample mean*, $\bar{x}$, is calculated by

$$
\bar{x} = \frac{x_1 + x_2 + \cdots + x_n}{n} = \frac{\sum^n_{i=1} x_i}{n}
$$

:::

. . .

::: {.dummy style="margin-top: 50px;"}
Calculate the *sample mean* for the following set of values:
:::

- 9, 9, 6, 3, 1 [&zigrarr; Sample mean = 5.6]{.printOnly}

## Sample Median

A *measure of centre* that is often coined as the "middle" value^[More formally known as the 50th percentile.] of a numeric variable. The (sample) median is one of five values that is always visualised by a **box plot**.

::: {.fragment .definition style="margin-top: 50px; width: 90%;" fragment-index=0}
Let $x_i$ be the $i$th value of a numeric variable and $n$ be the total number of values. The **sample median**, $m$, is calculated by

<ol>
  <li> Sorting the values in ascending (or descending) order. </li>
  <li class="fragment" data-fragment-index=1> <ul>
    <li class="fragment" data-fragment-index=1> If $n$ is **odd**: $m$ is the $\frac{n + 1}{2}$th sorted value</li>
    <li class="fragment" data-fragment-index=2> If $n$ is **even**: $m$ is the **sample mean** of the sorted values "above" and "below" $\frac{n + 1}{2}$</li>
  </ul> </li>
</ol>
:::

. . .

::: {.dummy style="margin-top: 50px;"}
Calculate the *sample median* for the following sets of values:
:::

<ul>
  <li> 7, 9, 3, 3, 10 [&zigrarr; Sample median = 7]{.printOnly}
  <li class="fragment"> 10, 0, 6, 3, 10, 6 [&zigrarr; Sample median = 6]{.printOnly}
</ul>


## Sample Standard Deviation

A *measure of spread* that summarises the variability of values from the **sample mean**.

. . .

::: {.definition style="margin-top: 50px; width: 90%;"}
Let $x_i$ be the $i$th value of a numeric variable and $n$ be the total number of values. The **sample standard deviation**, $s$, is calculated by

$$
\begin{aligned}
  s &= \sqrt{\frac{(x_1 - \bar{x})^2 + (x_2 - \bar{x})^2 + \cdots + (x_n - \bar{x})^2}{n - 1}} \\
    &= \cdots \\
    &= \sqrt{\frac{\sum^n_{i=1} (x_i)^2 - n\bar{x}^2}{n - 1}}
\end{aligned}
$$

:::

. . .

::: {.dummy style="margin-top: 50px;"}
Calculate the *sample standard deviation* for the following set of values:
:::

- 6, 8, 1, 5, 5 [&zigrarr; Sample standard deviation = 2.5495 (4 dp)]{.printOnly}

## [Ex 1]{.red} Sample Statistics in [Python]{.cornflowerblue}

:::: {.columns}
::: {.column width="49%"}
```{python Ex1-python}
#| echo: true

# Load in the pandas package
import pandas

# Read in the alloy.xlsx file
alloy_df = pandas.read_excel("datasets/alloy.xlsx")

# Calculate the sample mean
alloy_df["strength"].mean()

# Calculate the sample median
alloy_df["strength"].median()

# Calculate the sample standard deviation
alloy_df["strength"].std()
```
:::
::: {.column width="1%"}
:::
::: {.column}
```{r alloy-stackeddotplot}
#| fig-dpi: 300
#| fig-height: 4
#| fig-width: 7
#| out-width: 100%
#| fig-cap: "**Figure:** Distribution of 80 aluminium-lithium alloy specimens' [**Figure:**]{style=\"visibility: hidden;\"} compressive strength."
```
:::
::::

::: {.aside}
**Note:** [iNZight Lite](https://lite.docker.stat.auckland.ac.nz/) is a web app that can do most of the (exact) calculations *without code* for the Statistics component.
:::

## [Glossary]{.darkorange}

::::: {.columns}
:::: {.column}
::: {.parskip-16pt style="font-size: 0.8em; padding-right: 12pt;"}
**Centre**  
The location of a numeric variable's "typical" value, e.g. sample mean and median.

**Percentile**  
The location of a numeric variable's *n*th value after it is *sorted*, e.g. 50th percentile (sample median), 25th percentile (lower quartile), and 75th percentile (upper quartile).

**Spread**  
The variability of a numeric variable's values, typically from a *measure of centre*, e.g. sample standard deviation, range, and interquartile range.

**Outlier**  
A value of a numeric variable is notably distinct from all other values.

**Cluster**  
A distinct group of values.

**Sample size**  
Short for: "The total number of observations in a dataset".
:::
::::
:::: {.column}
::: {.parskip-16pt style="font-size: 0.8em; padding-left: 12pt;"}
**Shape**  
The *form* of a numeric variable's distribution in terms of the *observed frequencies* of values, e.g. U-shaped, triangle-shaped, and bell-shaped.

**Mode**  
The location of a numeric variable's most *frequent* value(s). There can be *more than* one. Furthermore, it does not always correspond to a *measure of centre*.

**Symmetrical**  
A distribution whose **shape** approximately matches when "folded" along a *measure of centre*.

**Skewed to the left**  
A distribution whose **shape** has values pile up on the *right*, and its tail extends relatively far out to the *left*.

**Skewed to the right**  
A distribution whose **shape** has values pile up on the *left*, and its tail extends relatively far out to the *right*.
:::
::::
:::::

## Practice Exercises {visibility="uncounted"}

:::: {.columns style="font-size: 0.8em;"}
::: {.column width="60%"}

<ol style="line-height: 1.1; padding-right: 24pt;">
  <li style="margin-bottom: 50pt;">
  What is an observation, variable, and value?
  </li>

  <li style="margin-bottom: 50pt;">
  Calculate the sample mean, sample median, and sample standard deviation for the following set of values:
    <ul>
      <li style="list-style-type: disc;">
      70, 63, 36, 67, 60, 64, 61
      </li>
    </ul>
  </li>

  <li>
  The `fairy-bay.csv` file describes a subset of data collected by Benjamin et al. (2022) about the green-lipped mussels sampled from Fairy Bay, New Zealand. For each mussel, they measured its shell length in millimetres ([**Length**]{.codeColour}) and categorised the type of plot ([**PlotType**]{.codeColour}) they were sampled from, either [*Soft Sediment*]{.codeColour} or [*Shell Substrate*]{.codeColour}.
    <ol>
      <li style="list-style-type: lower-alpha; line-height: 1.1;">
      Use [Python]{.cornflowerblue} (or other software) to open `fairy-bay.csv` file with the `pandas.read_csv()` function.
      </li>
      <li style="list-style-type: lower-alpha; line-height: 1.1;">
      Use [Python]{.cornflowerblue} to calculate the sample mean, sample median, and sample standard deviation of the mussel shell lengths.
      </li>
      <li style="list-style-type: lower-alpha; line-height: 1.1;">
      Use **Figure A** and the numeric summaries calculated in part b. to comment on features of the data.
      </li>
    </ol>
  </li>
</ol>

:::
::: {.column width="40%"}
```{r fairybay}
#| fig-dpi: 300
#| fig-height: 4
#| fig-width: 7
#| out-width: 100%
#| fig-cap: "**Figure A:** Distribution of 84 green-lipped mussels' shell lengths."

fairy.df <- read.csv("datasets/fairy-bay.csv")

ggplot(fairy.df, aes(x = Length)) +
  labs(x = "Shell Lengths (mm)", title = "Distribution of Green-Lipped Mussel Shell Lengths", y = NULL) +
  theme_bw() +
  geom_histogram(fill = "lightgreen", colour = "black", bins = 23) + 
  scale_y_continuous(breaks = NULL)
```
:::
::::

::: {.aside}
Benjamin, E. D., Hillman, J. R., Handley, S. J., Toone, T. A., & Jeffs, A. (2022). The Effectiveness
of Providing Shell Substrate for the Restoration of Adult Mussel Reefs. *Sustainability*, **14**(23),
15746.
:::
