---
title: "Inference for a Difference in Means"
author: "ENGEN102-23G (HAM) - Engineering Maths and Modelling 1B"
format:
  revealjs: 
    slide-number: c/t
    progress: true
    auto-stretch: false
    height: 1080
    width: 1920
    theme: reveal-extra/quartz.scss
    template-partials:
      - reveal-extra/title-block.html
    highlight-style: breeze
    code-line-numbers: false
    embed-resources: false
editor: source
---
  
```{r setup}
#| include: false
library(knitr)
library(ggplot2)
```

## Motivation

## Sampling Distribution of the Difference in Sample Means

## Standard Error of *x&#772;*~1~ &minus; *x&#772;*~2~

## [Ex ?:]{.red} Handedness and Earnings

::: {.aside}
Ruebeck, C. S., Harrington Jr, J. E., & Moffitt, R. (2007). Handedness and earnings. *Laterality*, **12**(2), 101--120.
:::

## Estimation of the Difference in Means

## Interpretation of the Confidence Interval

## Test for the Difference in Means

## Using Student's *t*-distribution to Calculate the *p*-value

## Interpretation of the Hypothesis Test

## [Ex ?:]{.red} Jet Noise

::::: {.columns}
:::: {.column width="70%"}
<blockquote style="border-left: 0.25em solid #685f5f;">
The `jet-noise.csv` file describes a study investigated the noise levels of a random sample of wide-bodied jets, and narrow-bodied jets were measured immediately after their wheels left the ground at an international airport. Why noise levels? The researchers thought it may influence the frequency of birds being hit by aircraft (as it is regarded as a safety hazard).
</blockquote>

::: {.fragment fragment-index=1 .orchid-table style="font-size: 0.9em; margin-top: 50px; width: 90%;"}
+:----------+:----------------------------------------------------------------------------------------------------------------------------+
| **noise** | A number that denotes the compressive strength of an aluminium-lithium alloy, measured in decibels (dB). 
+-----------+-----------------------------------------------------------------------------------------------------------------------------+
| **body**  | A number that denotes the compressive strength of an aluminium-lithium alloy, measured in decibels (dB). 
+-----------+-----------------------------------------------------------------------------------------------------------------------------+
:::
::::
:::: {.column width="30%" .fragment fragment-index=0}
```{r alloy} 
#| results: asis
#| classes: excel

noise.df <- read.csv("datasets/jet-noise.csv")
 
rbind(colnames(noise.df), head(noise.df, n = 14), rep("...", ncol(noise.df))) |>
  kable(format = "html", row.names = TRUE, col.names = LETTERS[1:ncol(noise.df)])
```
::::
:::::

::: {.aside}
Data from Wild, C. J., & Seber, G. A. F. (2000). *Chance Encounters: A First Course in Data Analysis and Inference*. Wiley.
:::

##

## Assumptions for Inference on the Difference in Means

## Practice Exercises {visibility="uncounted"}
